The Concurrent Calculi Formalisation Benchmark

### 2024-02-01 short-paper/short-paper.tex
Add Alberto's comments.
Co-authored-by: Alberto Momigliano <momigliano@di.unimi.it>

\subsubsection{General design remarks.}
Like the POPLMark and POPLMark Reloaded challenges, our challenge is not meant to be comprehensive:
applications such as multiparty session types~\cite{10.1145/2827695,10.1145/3290343}, choreographies~\cite{DBLP:journals/jar/CruzFilipeMP23}, conversation types~\cite{DBLP:journals/tcs/CairesV10}, psi-calculi~\cite{lmcs:696}, or encodings between different calculi~\cite{DBLP:journals/iandc/Gorla10,DBLP:conf/forte/CairesP16,DBLP:journals/iandc/DardhaGS17,DBLP:conf/ecoop/ScalasDHY17,DBLP:journals/iandc/KouzapasPY19,10.1145/3479394.3479407} are not directly covered.  Still, these (and other) applications need the basic techniques that our challenge problems emphasize:
as per design goal \ref{item:goal-reusability},
our problems are drawn from the basic meta-theory of process calculi (without requiring the development of new theory), and focus on well-known theorems
and results that involve interesting proof techniques that may be reused in further work.
To achieve design goal \ref{item:goal-comperison-accessibility},
we have formulated our challenge problems to explore the three aspects (linearity, scope extrusion, coinduction) independently, so that they may be solved individually and in any order;
each problem should be reasonably small and easily understood with basic knowledge of textbook concurrency theory, process calculi, and type theory.  The process calculus used in each challenge
focuses on the features that we want to emphasize, and omits all constructs
(such as choices) that would complicate the mechanisation without bringing additional tangible
benefits and insights. 
The minimality and uniformity of the calculi also allows us to target design
goal \ref{item:goal-tutorials}.

@@

\subsubsection{Linearity.}
For our challenge on linear reasoning we have chosen a type preservation theorem
for a session type system, which allows us to introduce linear
reasoning with few definitions. Accordingly, we have chosen not to
include channel delegation and to use a reduction instead of a
transition system semantics.
Linearity is of course interesting in many other sub-fields of programming languages, but we have chosen to base our challenge on a session type system primarily due to the common use of session types in concurrency theory.
Inspired by Vasconcelos~\cite{Vasconcelos2012}, we use a syntax where
restriction binds two names. This presentation makes the duality
of the two names obvious in the type system.
We have chosen a simple notion of well-formedness for our challenge,
and do not, for instance, consider deadlocks. While more
sophisticated notions of well-formedness are interesting, the
proofs of these properties would complicate the challenge, distracting from the linear aspects.

\subsubsection{Scope extrusion.}
This is the challenge most closely related to the POPLMark
challenge since it concerns the properties of binders in restrictions.
It is not obvious that the techniques used to solve the original challenges about binders can be easily extended to reasoning about scope extrusion.
For example, techniques like higher-order abstract syntax, or well-scoped
de\ Bruijn indices do not have the concept of free variables and thus encoding this challenge with them requires some attention.

@@

Our coinduction challenge concerns strong barbed bisimulation and congruence.
Though weak barbed congruence is a more common behavioural equivalence for \(\pi\)-terms, here we prefer strong equivalences to avoid the need to abstract over the number of internal transitions, % when considering the strong notions of equivalence,
thus simplifying the theory.
We have chosen not to include delegation in the coinduction challenge since it is orthogonal to our primary aim of exploring coinductive proof techniques.

>>>>

\subsubsection{General design remarks.}
Like the POPLMark and POPLMark Reloaded challenges, our challenge is not meant to be comprehensive:
applications such as multiparty session types~\cite{10.1145/2827695,10.1145/3290343}, choreographies~\cite{DBLP:journals/jar/CruzFilipeMP23}, conversation types~\cite{DBLP:journals/tcs/CairesV10}, psi-calculi~\cite{lmcs:696}, or encodings between different calculi~\cite{DBLP:journals/iandc/Gorla10,DBLP:conf/forte/CairesP16,DBLP:journals/iandc/DardhaGS17,DBLP:conf/ecoop/ScalasDHY17,DBLP:journals/iandc/KouzapasPY19,10.1145/3479394.3479407} are not directly covered.\alb{Do we need all those citations?}  Still, these (and other) applications need the basic techniques that our challenge problems emphasize:
as per design goal \ref{item:goal-reusability},
our problems are drawn from the basic meta-theory of process calculi (without requiring the development of new theory), and focus on well-known theorems
and results that involve interesting proof techniques that may be reused in further work.
To achieve design goal \ref{item:goal-comperison-accessibility},
we have formulated our challenge problems to explore the three aspects (linearity, scope extrusion, coinduction) independently, so that they may be solved individually and in any order;
each problem should be reasonably small and easily understood with basic knowledge of textbook concurrency theory, process calculi, and type theory.  The process calculus used in each challenge
focuses on the features that we want to emphasize, and omits all constructs
(such as choices) that would complicate the mechanisation without bringing additional tangible
benefits and insights. 
The minimality and uniformity of the calculi also allows us to target design
goal \ref{item:goal-tutorials}.

@@

\subsubsection{Linearity.}
For our challenge on linear reasoning we have chosen a type preservation theorem
for a session type system, which allows us to introduce linear
reasoning with few definitions. Accordingly, we have chosen not to
include channel delegation and to use a reduction instead of a
transition system semantics.
Linearity is of course interesting in many other sub-fields of programming languages, but we have chosen to base our challenge on a session type system primarily due to the common use of session types in concurrency theory.\alb{This is weak and somewhat tautological, as a motivation}
Inspired by Vasconcelos~\cite{Vasconcelos2012}, we use a syntax where
restriction binds two names. This presentation makes the duality
of the two names obvious in the type system.
We have chosen a simple notion of well-formedness for our challenge,
and do not, for instance, consider deadlocks. While more
sophisticated notions of well-formedness are interesting, the
proofs of these properties would complicate the challenge, distracting from the linear aspects.

\subsubsection{Scope extrusion.}
This is the challenge most closely related to the POPLMark
challenge since it concerns the properties of binders in restrictions.\alb{I would argue also wrt input}
It is not obvious that the techniques used to solve the original challenges about binders can be easily extended to reasoning about scope extrusion.\alb{I would say not obvious what the pros and cons are, since in fact those techniques (DB, locally nameless, HOAS, nominal) have been used.}
For example, techniques like higher-order abstract syntax, or well-scoped
de\ Bruijn indices do not have the concept of free variables and thus encoding this challenge with them requires some attention.

@@

Our coinduction challenge concerns strong barbed bisimulation and congruence.\alb{we motivate why strong vs weak, but not why barbed}
Though weak barbed congruence is a more common behavioural equivalence for \(\pi\)-terms,\alb{? why terms?} here we prefer strong equivalences to avoid the need to abstract over the number of internal transitions, % when considering the strong notions of equivalence,
thus simplifying the theory.
We have chosen not to include delegation in the coinduction challenge since it is orthogonal to our primary aim of exploring coinductive proof techniques.
